\Large\textsc{Regression And Resampling Techniques \\ FYS-STK4155 \\ Assignment One}

### 2019-09-18 assignment_one/article/main.tex
Added terrain.pdf image
Added optional X-input for validation prediction!

\documentclass[dvipsnames, article, a4paper, oneside, 12pt]{memoir}
\linespread{1.25}
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\newcommand{\var}{\mathbb{V}}
\newcommand{\cov}{\mathrm{Cov}}
\newcommand{\OLS}{\mathrm{OLS}}
\newcommand{\SD}{\mathrm{SD}}
\newcommand{\ridge}{\mathrm{ridge}}
\newcommand{\lasso}{\mathrm{lasso}}
\newcommand{\python}{\textsf{Python}{}}
\newcommand{\sgn}{\,\mathrm{sign}{}}

@@

  We sample the function \( F \) both with and without noise at \( N = 5000 \) points in each
  parameter direction, yielding a matrix of function values
  \begin{align}
    \begin{split}
      \mat{Z} &= \left[F(x_i, y_j) \right]_{ij}, \\
      \mat{Z}^\star &= \left[F(x_i, y_j) + \varepsilon_{ij}\right]_{ij},
    \end{split}
  \end{align}
  where each \( \varepsilon_{ij} \sim \N(0, \sigma^2) \). The design matrix \(
  \mat{X} \) is computed using the \textsc{PolynomialFeatures}-method from the
  \textsc{sklearn} package. The dimension of the bi-quintic polynomial space is
  21, yielding a design matrix of size \( N^2 \times 21 \).

  For the real data set, we use the supplied \textsc{Tiff} data of an area
  close to Stavanger in Norway, as displayed in \cref{fig:terrain}. The
  data-set consists of approximately \( 1800 \times 3600 \) data-points. 
  \begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{images/terrain.png}
    \caption{Real terrain data from a region close to Stavanger, Norway. For
    the regression tasks, we select data from small rectangular regions to keep
    the data-sizes manageable. The data is supplied in the file
    \texttt{SRTM\_data\_1\_Norway.tif}.}
  \label{fig:terrain}
  \end{figure}

@@

  We are interested in computing certain statistical quantities pertaining to
  our data and corresponding model function. Firstly, the \emph{variance} of
  the model parameters \( \beta \) over all possible data sets tells us how
  sensitive \( \beta \) is to changes in the sampled data set. Using
  \cref{eq:optimal} for the optimal OLS-parameters and setting \( \mat{A} =
  (\X^T\X)^{-1}\X^T\), we compute:
  \begin{align}
	  \label{eq:ols_variance}
	  \cov\left[\beta_{\OLS}\right] &= \cov\left[\mat{A}\y\right] = \mat{A} \cov\left[\y\right] \mat{A}^T = \mat{A}\mat{A}^T\sigma^2 = (\X^T{\X})^{-1} \sigma^2.
  \end{align}
  In this calculation, we have used \cref{as:general_linear} to deduce that \(
  \cov[ \y ] = \diag{\sigma}\), a diagonal matrix, which commutes with other
  matrices. It then follows that the estimated variance of \( \beta_{\OLS} \) is given by
  \begin{equation}
	  \var[\beta_{\OLS}] = [\sigma^2 (\X^T\X)^{-1}_{ii} ]_{i=1}^n.
  \end{equation}
  Furthermore, by direct computation, we see that \( \beta_{\OLS} \) is an \emph{unbiased} estimator,
  as 
  \begin{equation}
	  \expect[\bias_{\OLS}] = \beta.
  \end{equation}
  With both the expected value and the variance of \( \beta \), we may
  formulate a confidence interval for \( \beta \). Thus for an estimated
  parameter \( \beta_i \) we construct the
  \(95\%\)-confidence interval as
  \begin{align}
	  \begin{split}
		  I_i &= (\expect[\beta_i] - 1.96 \times \SD[\beta_i], \expect[\beta_i] + 1.96 \times \SD[\beta_i]) \\
		      &= \left(\beta_i - 1.96 \times \sigma \sqrt{(\X^T\X)^{-1}_{ii}}, \beta_i + 1.96 \times \sigma \sqrt{(\X^T\X)^{-1}_{ii}}\right).
	  \end{split}
  \end{align}
  Here \( \SD[\beta_i] \) denotes the standard deviation of \( \beta_i \).  The
  interpretation of this interval is  ``under repeated experiments, the number
  of computed confidence intervals that actually include the true parameter,
  will tend to \( 95\% \)''.

@@

@@

  \section{Terrain Data}

>>>>

\documentclass[dvipsnames, article, a4paper, oneside, 12pt]{memoir}
\newsubfloat{figure}
\linespread{1.25}
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

@@

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}
\newcommand{\diag}[1]{\mathrm{diag}(#1)}
\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
\newcommand{\R}[1]{\mathrm{R}^2(#1)}
\newcommand{\expect}{\mathbb{E}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}
\newcommand{\OLS}{\mathrm{OLS}}
\DeclareMathOperator{\SD}{SD}
\newcommand{\ridge}{\mathrm{ridge}}
\newcommand{\lasso}{\mathrm{lasso}}
\newcommand{\python}{\textsf{Python}{}}
\newcommand{\sgn}{\,\mathrm{sign}{}}

@@

  We sample the function \( F \) both with and without noise at \( N = 5000 \) points in each
  parameter direction, yielding a matrix of function values
  \begin{align}
    \begin{split}
      \mat{Z} &= \left[F(x_i, y_j) \right]_{ij}, \\
      \mat{Z}^\star &= \left[F(x_i, y_j) + \alpha\varepsilon_{ij}\right]_{ij},
    \end{split}
  \end{align}
  where each \( \varepsilon_{ij} \sim \N(0, \sigma^2) \) and \( \alpha \) is
  the \emph{signal-to-noise ratio}. The design matrix \( \mat{X} \) is computed
  using the \textsc{PolynomialFeatures}-method from the \textsc{sklearn}
  package. The dimension of the bi-quintic polynomial space is 21, yielding a
  design matrix of size \( N^2 \times 21 \).

  For the real data set, we use the supplied \textsc{Tiff} data of an area
  close to Stavanger in Norway, as displayed in \cref{fig:terrain}. The
  data-set consists of approximately \( 1800 \times 3600 \) data-points. 
  \begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{images/terrain.pdf}
    \caption{Real terrain data from a region close to Stavanger, Norway. For
    the regression tasks, we select data from small rectangular regions to keep
    the data-sizes manageable. The data is supplied in the file
    \texttt{SRTM\_data\_1\_Norway.tif}.}
  \label{fig:terrain}
  \end{figure}

@@

  We are interested in computing certain statistical quantities pertaining to
  our data and corresponding model function. Firstly, the \emph{variance} of
  the model parameters \( \beta \) over all possible data sets tells us how
  sensitive \( \beta \) is to changes in the sampled data set. Using
  \cref{eq:optimal} for the optimal OLS-parameters and setting \( \mat{A} =
  (\X^T\X)^{-1}\X^T\), we compute:
  \begin{align}
	  \label{eq:ols_variance}
	  \cov\left[\beta_{\OLS}\right] &= \cov\left[\mat{A}\y\right] = \mat{A} \cov\left[\y\right] \mat{A}^T = \mat{A}\mat{A}^T\sigma^2 = (\X^T{\X})^{-1} \sigma^2.
  \end{align}
  In this calculation, we have used \cref{as:general_linear} to deduce that \(
  \cov[ \y ] = \diag{\sigma}\), a diagonal matrix, which commutes with other
  matrices. It then follows that the estimated variance of \( \beta_{\OLS} \) is given by
  \begin{equation}
	  \var[\beta_{\OLS}] = [\sigma^2 (\X^T\X)^{-1}_{ii} ]_{i=1}^n.
  \end{equation}
  Furthermore, by direct computation, we see that \( \beta_{\OLS} \) is an \emph{unbiased} estimator,
  as 
  \begin{equation}
	  \expect[\beta_{\OLS}] = \beta.
  \end{equation}
  With both the expected value and the variance of \( \beta \), we may
  formulate a confidence interval for \( \beta \). Thus for an estimated
  parameter \( \beta_i \) we construct the
  \(95\%\)-confidence interval as
  \begin{align}
	  \begin{split}
		  I_i &= (\expect[\beta_i] - 1.96 \times \SD[\beta_i], \expect[\beta_i] + 1.96 \times \SD[\beta_i]) \\
		      &= \left(\beta_i - 1.96 \times \sigma \sqrt{(\X^T\X)^{-1}_{ii}}, \beta_i + 1.96 \times \sigma \sqrt{(\X^T\X)^{-1}_{ii}}\right).
	  \end{split}
  \end{align}
  Here \( \SD[\beta_i] \) denotes the standard deviation of \( \beta_i \).  The
  interpretation of this interval is  ``under repeated experiments, the number
  of computed confidence intervals that actually include the true parameter,
  will tend to \( 95\% \)''.

@@

  In this section we present some of the results of the numerical simulations.
  We start simple by first considering the effects of model complexity --- in
  terms of polynomial degree --- using OLS regression on the Franke dataset. We
  assess the model accuracy by means of the mean squared error and the \(
  R^2\)-score discussed previously. In order to increase the reliability of the
  statistical estimates, we employ \( k\)-fold cross validation.

  We then take a closer look at Ridge and Lasso regression techniques, and find
  the optimal hyper-parameter \( \lambda \). Again, we use the Franke dataset.
  In this context, we also see the effects of the bias-variance tradeoff.

  Finally, we compare our methods on real world data, namely the terrain data
  in \cref{fig:terrain}.

@@

  \begin{figure}
    \centering
    \subbottom{%
      \includegraphics[width=0.45\linewidth]{images/OLS_MSE_score.pdf}
    }
    \subbottom{%
      \includegraphics[width=0.45\linewidth]{images/OLS_R2_score.pdf}
    }
    \caption{Ordinary least squares regression on samples from the Franke
      function. The mean squared error (left) and the \(R^2\)-score (right) as
      a function of the polynomial degree \( d \). The values are displayed
      both for the true sample data (solid) and the noisy samples (dashed). The
      noise is distributed as \( \N(0, \tfrac{5}{100})\).}
    \label{fig:ols_scores}
  \end{figure}

  \subsubsection{Ordinary Least Squares}

  We sample \( N = 1000 \) points in each parameter direction and compute the
  corresponding \( N^2 \) funciton values of the Franke function both with and
  without added noise. The noise was selected to be distributed as \( \N(0,
  1) \) with a signal-to-noise ratio \( \alpha = 0.05\).

  We then perform ordinary least squares, and compute the mean squared error as
  well as the \( R^2 \)-score for both the true and the noisy data. These are
  presented in \cref{fig:ols_scores}. With the Franke function being a weighted
  sum of exponentials, the MSE-error tends to zero as the polynomial degree
  increases. The added noise reduces the convergence rate, as is to be
  expected.

  In addition to the evaluation metrics, we also compute confidence intervals
  for each model parameter. These are displayed in \cref{tab:ols_confidence}
  for each polynomial degree up to and including \( d = 5 \).

  \begin{table}
    \caption{}
    \label{tab:ols_confidence}
  \end{table}
  \section{Terrain Data}
